---
title: Moonlander ZSA Ergonomic Keyboard -- Review
author: Alec Wong
cover: https://www.zsa.io/static/8567515c1ffbc03f83c7bfa2b29bd155/4ad1b/1-hero-white.png
date: '2020-12-09'
slug: moonlander-zsa-review
categories: [equipment]
tags: [moonlander, zsa, mechanical keyboards, ergonomics]
---

```{r setup, include = FALSE}
library(dplyr)
library(data.table)
library(ggplot2)

knitr::opts_knit$set(root.dir = here::here())
```

# Moonlander ZSA Review

I bought this thing in late October, a few days after I'd had some pain in my
wrist at work. This thing comes direct from Taiwan (free shipping!), and the
wait was rather agonizing, checking the tracking every 10 minutes or so. At
$375, this is not a budget keyboard by any means, but I figured the health
benefits avoiding wrist injury was worth the price tag. Reviews indicated that
the build quality is extremely high, so I went for it.

The TL;DR is that this is a forward-thinking, excellently-devised split
keyboard, able to keep up with all of your configuration needs. The tenting and
split design offers the advertised ergonomic benefits, but I find the
ortholinear layout less appropriate for gaming. I'm still using my old keyboard
for work until my confidence increases with the Moonlander. Replace the keycaps
with sculpted keycaps ASAP.

## Typing speed

As expected, typing speed ***tanked*** in the first week that I worked with this
keyboard. The manufacturer ZSA provides a terrific app called Oryx that allows
you to configure your layout, test your typing speed on text or code, and
monitor the keystrokes live -- this is what I used primarily to assess my typing
speed. 

```{r}
datasets     = list.files('static/assets/data/')
paths        = list.files('static/assets/data/', full.names = TRUE)
dates        = substr(datasets, start = 1, stop = 10) %>% as.Date()
last_dataset = paths[which.max(dates)]
typingData   = fread(last_dataset)

background_color = rgb(31/255, 34/255, 42/255)

typingData_long = typingData %>% 
  tidyr::pivot_longer(cols = c("Accuracy", "WPM"), names_to = "Metric", values_to = "Value")

ggplot(typingData_long) + 
  aes(x = Date, y = Value, color = Type) +
  geom_point() + 
  geom_smooth(se = FALSE, formula = y ~ x) + 
  facet_wrap(~Metric, scales = 'free') + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_bw() + 
  theme(
    panel.background = element_rect(fill = background_color),
    plot.background  = element_rect(fill = background_color),
    strip.background = element_rect(fill = background_color),
    strip.text = element_text(color = 'gray90'),
    legend.background = element_rect(fill = background_color),
    legend.key = element_rect(fill = background_color),
    text = element_text(color = 'gray90'),
    axis.text = element_text(color = 'gray90'),
    panel.grid = element_line(color = 'gray20')
  )
```

My typing speed dropped to as low as ~20 words per minute for code in the first
few times typing on the Moonlander. This should be expected; the ortholinear,
columnar layout takes time to get used to, and the location of symbols did not
correspond to any existing muscle memory on QWERTY layouts.

However, over time, my speed has improved; I've actually exceeded my typing
speed on an ordinary layout (which was previously typically around 85 WPM)
reaching as high as `r max(typingData$WPM)` WPM. This is certainly a welcome
change.

Typing code is going to be slower due to the symbols accessed, but familiarity
with my layout and keeping a layout as intuitive as possible to existing muscle
memory returned my speed to roughly 60 WPM. Accuracy is fairly constant, around
96% for both types of typing (Oryx does not register data points for results <
95%).

## Layouts

Speaking of layouts, [my
layout](https://configure.ergodox-ez.com/moonlander/layouts/q9AGA/latest/0)
changed drastically and often compared to the default Moonlander. The bit that
gave me the most trouble was where to keep the symbols:

* Should I keep them in a dedicated layer? 
* Put them on the left-hand (like the default layout has it), right-hand, or both sides?
* Which symbols should exist on the base layer?

Ultimately, I decided that symbols should all mostly live in their own layer,
but the muscle memory from an ordinary layout should not be forsaken unless
there was a good speed, convenience, or ergonomic reason.

I used the key where caps-lock normally goes to access the symbols layer, and
moved all of the symbols from the number keys down to the home row (for the most
part). This way, existing muscle memory needs not be reprogrammed too far.

At this moment, I have 8 layers;

* Base
* Arrows (accessed from the semi-colon, like the base layer)
* Numpad
* Symbols
* Gaming
* Colors
* Control (media and more arrow keys)
* R/IDE 

## Ergonomics

Perhaps it's due to the unfamiliar layout, but I do continue to feel some strain
on my hand tendons & muscles. I may go so far as to say that the ortholinear
layout does not achieve its ergonomic purpose, and causes some strain compared
to the staggered key layout. At this moment, tension in my hands is higher, and
some pain flickers in my right-hand-side ring finger tendons. I wonder how much
of this is due to the combination of the pinky button activating the "Arrows"
layer for quick arrow navigation while typing.

I appreciate the tenting and the split layout, and I do think this does
alleviate discomfort. Normal keyboards feel somewhat cramped in comparison. I
find myself typing in the recommended way, arms at a 90° angle, and hands
floating above the keys instead of resting on the wrist-wrest.

The keycaps are the same size as my old board, but keys feel *smaller* on a
staggered layout keyboard. I guess this implies that I feel like my fingers are
traveling farther on average to hit keys with the Moonlander. I think that there
is some merit to this; the staggered layout has keys oriented in an _almost_
hexagonal arrangement; while some keys are farther to reach on a standard
keyboard ("B" is one instance) I think for the most part the ortholinear layout
incurs more travel distance as the grid arrangement is less space efficient than
the hexagonal arrangement.

Compare the distance between "F" and "6"; on a standard layout this is about `r 37.5 - 33`
centimeters. On the ortholinear layout, it is `r 22 - 17`cm to the key to the
right of "5". This is a difference of half a centimeter; this sounds
insignificant but it is 11% farther to travel!

The ortholinear layout also requires some more contortion in certain contexts.
For example; take the position of the letter "F": on an ordinary staggered
layout, my finger can comfortably press R, T, G, V, C, and B without adjusting
the rest of my left hand. On the ortholinear layout, I lose access to the letter
"C"; I must use a contortion to hit the letter with my index finger that
disrupts the position of my other fingers.

One may say that I should instead learn to press "C" with my middle finger;
perhaps this is proper, but I have developed my typing style with an underlying
principle of mitigating repositioning of fingers; I feel like it is most
efficient to *not* have to move one's hands too much when touch-typing. Pressing
the center of the "C" key using the middle finger on a standard layout requires
lifting the index finger out of the way; otherwise the middle finger intersects
the index. Better to use the index to press "C", as it requires less
readjustment, in my opinion.

This is about the only concern for normal typing, but the grid vs. hexagonal
arrangement has a greater bearing in gaming.

## Gaming

Lots of buttons are available by the left hand between the additional row on the
bottom and the thumb-cluster keys. This is certainly a plus, although I find the
accuracy of the thumb-cluster keys to be lower in general, so they are reserved
for less critical actions, usually reserved by keys "B", "N", and "M". This is
a *very* nice improvement, not having to reach all the way over to the "M" key
with the left hand, which is often used to open up a map.

The major concern about the grid layout for gaming is due to what was mentioned
in the previous section; there is more repositioning that needs to be done due
to the grid arrangement.

Take the common gaming situation where digits 1-4 access certain inventory
slots, and WASD are for movement. On a standard keyboard, while "W" is being
held by the middle finger (you're in the process of moving your character
forward), the digits "2" and "3" are accessible to be pressed by the ring and
index fingers respectively, because they are staggered up and to the left and
right of "W". In fact, it's not difficult to press all three simultaneously.

With the columnar layout, having "W" pressed by the middle finger results in
much greater contortion to access "2" in particular. "3" is still accessible,
but it's much too far high now due to the vertical shifting of the layout. Try
to press all three simultaneously; it's not really easy to do without hitting
other keys by mistake, and not without repositioning the middle-finger on "W".
Add left-shift for sprinting, and you now can't really press "2" at all.

"X" is the same way; while "W" is held down, "X" can be easily pressed using the
index finger with a staggered layout, but more movement of the hand is required
with the columnar layout; the hand needs to rotate nearly 45°. 

Given that equipment swaps typically need to be quick in response to changing
circumstances, I would trade the extra thumb buttons and bottom row for the
easier-to-access "2" and "X" buttons of the staggered layout. It _is_ nicer
having WASD in a grid, however.

I haven't experimented with remapping everything to the right one (W to E, A to
S, D to F, etc.); I could see this fixing the distance traveled to hit "3", but
I don't see it fixing the "2" and "X" issues.

I moved shift to where CAPS was, and left-control to where shift normally is.
This results in a more neutral position for the hands, which is admittedly nice.
Space is the key below "V".

## Programming

At this time, despite the multitude of layers and configurations, I still feel
more productive on a full layout. Why? I just feel like having each symbol
mapped to one button is more efficient than multiple layers per key. Call me
simple, but it's faster (fewer key strokes) and it's an easier mental model.
Perhaps I'll feel differently in a few months, when I've mastered this layout.

## Build

I think the quality of the build is fantastic. The lighting, materials,
software, and everything else are top-tier.

I replaced the original keycaps with HyperX sculpted keycaps (the pudding
variety); I was pressing wrong keys (particularly with the numeric keys) way too
often due to the lack of vertical differentiation between farther keys. Travel
is _far_ in some cases; take "5": the distance on a normal keyboard replicated
on the ortholinear layout brings me to the lower right corner of "4". An
additional few millimeters of travel are required to hit the "5" key, making it
all the more important to have some vertical profile to avoid hitting "T" by
accident. Sculpted keys made a tremendous difference in my quality of life using
this board, I highly recommend swapping them out.

One gripe I have is the lack of grip; when the keyboard is in a flat position,
the underbody grips contact the surface and prevent the device from slipping.
When the keyboard is in the tented position, however, the grips no longer
contact the surface and the keyboard slides around way too much. I fixed this
with a washcloth underneath the left-hand component, but there should be
friction contacts on the wrist-rest and the thumb-cluster to avoid this sliding
problem.

Last, the wrist-rests are somewhat rattly when weight is shifted between the
left and right side of the rests. Most of the time this isn't an issue because
my hands are floating above the keys instead of on the rest. This will be less
of a concern going forward, as I've adopted the extra comfortable HyperX
gel-wrist rest.